---
title: "DZWS_karting"
author: "rnvdv"
date: "Monday, May 04, 2015"
output:
  html_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 10
    highlight: monochrome
    keep_md: yes
    number_sections: yes
    theme: flatly
    toc: yes
---

```{r, echo=FALSE, warning=FALSE,error=FALSE,results='hide',}
library(knitr)    # for fancy table support
library(ggplot2)  # fancy graphics
library(reshape2) # easy modification
library(car)      # that recode will come in handy!
library(texreg,quietly = T)   # fancy regression tables & plots

# make sure tables format correctly for github
options(knitr.table.format = "pandoc")

# Let's load some datas!
finals   <- read.table("final.txt",sep="\t",header = T, dec=",")
Q1       <- read.table("Q1.txt",sep="\t",header = T, dec=",")
Q2       <- read.table("Q2.txt",sep="\t",header = T, dec=",")
Q3       <- read.table("Q3.txt",sep="\t",header = T, dec=",")
training <- read.table("training.txt",sep="\t",header = T, dec=",")

start_and_end <- merge(finals,training,by="nr",suffixes = c(".start",".finals"))
```


# So we went karting....

'T was one of Daan's last days as a bachelor, and he decided he had little to lose: Let's go karting!

But who won, and why? Let's use this as a quick-'n-dirty presentation of R-markdown, the wonderous world of linear modelling and how people learn how to go-karting!

## Let's start with the end: the winners

We'll get it over with: Daan didn't quite win (and neither did I). So who did?

```{r}
kable(finals[c("pos","name","avg")], format="pandoc")
finals$name <- factor(finals$name,levels=(finals$name[order(finals$pos)]),ordered = T,)
ggplot(finals,aes(x=name,y=avg))+stat_identity(geom="bar")

```

That is a nice little show of the endgame, but what got us there? It's not the `r round(with(start_and_end, cor(pos.start,pos.finals)),digits=3)` correlation between our training positions and end positions. So what happened?

# What happened

If we look at the positions starting from our training rounds and leading up to the finals, we get quite a competitive field. Especially between the third qualification and the Finals, things seem to be changing rapidly. 

```{r,echo=FALSE}
# A quick function to make one-data.frame out of our seperate tables (complicated by their inconsistent columns)
superbind <- function(x)
  {
  i<-1
  all <- data.frame()
  for (df in x)
    {
    df$df_num <- i
    df <- data.frame(df)
    if (i==1){all <- rbind(all,df)}
    i <- i+1
    for (n in names(df)[!names(df)%in%names(all)]){all[[n]]<-NA}
    for (n in names(all)[!names(all)%in%names(df)]){df[[n]]<-NA}
    all <- rbind(all,df)
    
    }
  all
  }

kartdata <- superbind(list(training,Q1,Q2,Q3,finals))
kartdata$df_num <- recode(kartdata$df_num,"1='Training';2='Q1';3='Q2';4='Q3';5='Finals'",as.factor.result = T)
kartdata$df_num <- factor(kartdata$df_num, levels=levels(kartdata$df_num)[c(5,2:4,1)])
ggplot(kartdata,aes(x=df_num,y=pos,colour=name))+
  geom_line(aes(group=name),size=3)+
  geom_point(line="white",size=6)+
  scale_y_reverse()+# inverted y-axis to be more intuitive (you want being on top to be on top right? ) 
  xlab("round")+    # x-axis label
  ylab("position")+ # y-axis label
  ggtitle("Positions throughout the rounds") # graph title


```

## Personal improvement

If we look at it from the perspective of personal improvement, what does that give us?

```{r}

ggplot(kartdata,aes(x=df_num,y=avg, colour=name))+
  geom_line(aes(group=name),size=3)+
  geom_point(line="white",size=6)+
  scale_y_reverse()+xlab("round")+ ylab("inverted time")+ggtitle("Personal round averages throughout the rounds")

wide_kart <- reshape(data = kartdata,direction="wide", timevar="df_num",idvar="name")
wide_kart$improvement <- wide_kart$avg.Training - wide_kart$avg.Finals

kable(wide_kart[order(wide_kart$improvement,decreasing=T),c("name","improvement","avg.Training" ,"avg.Finals","pos.Training","pos.Finals")],caption = "Time improvement between training and round 10 averages")
```

So did the best learner win? It's not quite clear from that little diddy. So we'll throw it through a simple lm, just for the heck of it. But wait, how do the extra rounds figure into this (if your quick, you can do more rounds...)

```{r}
roundnames <- character()
for(n in paste(".",unique(kartdata$df_num),sep="")) for (ni in paste("X",1:15,n,sep="")) if (ni %in% names(wide_kart)) roundnames <- c(roundnames,ni)
wide_kart$rounds <- apply(wide_kart[roundnames],1,function(x){sum(!is.na(x))})

m <- lm(avg.Finals~avg.Training*rounds,wide_kart)
summary(m)
plotreg(m)

```

# The End (?)

That's it: the better your start, the more rounds you do, the better you end! 
Now theres a first-mover advantage for you!